Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-04-07 main_allen.tex
Partially done merging Allen's changes

\section{Introduction} 
$\ell_1$-minimization ($\ell_1$-min) has received much attention in recent
years due to important applications in compressive sensing
\cite{BrucksteinA2007} and sparse representation \cite{WrightJ2010-PIEEE}.  
$\ell_1$-min refers to finding the minimum $\ell$-norm solution to an
underdetermined linear system $\bb=A\xx$:
\begin{equation}
\min \|\xx\|_1\quad \mbox{ subj. to }\quad \bb = A\xx.
\label{eq:l1min}
\end{equation}
It is now well established that, under certain conditions
\cite{CandesE2005-IT_1,DonohoD2004}, the minimum $\ell_1$-norm solution is also
the \emph{sparsest} solution to the system \eqref{eq:l1min}.

@@

In this paper, we study parallelization of $\ell$-min on many-core CPU and GPU
architectures. Although $\ell_1$-min \eqref{eq:l1min} is a convex
program, conventional algorithms such as interior-point methods
\cite{ChenS2001-SIAM,TibshiraniR1996} are known to suffer from poor scalability
for large-scale applications such as face recognition. Recently, a number of
accelerated algorithms have been proposed that explicitly take advantage of
the special structure of $\ell_1$-min problems
\cite{LorisI2009,YangA2010-ICIP}. We investigate parallelization of a
state-of-the-art $\ell_1$-min solution based on the classical framework of
\emph{augmented Lagrange multiplier} (ALM) \cite{BertsekasD2003,YangA2010-ICIP}.

While solving \eqref{eq:l1min_denoise} can achieve exceptional
recognition accuracy on public databases, its success relies on the assumption
that the query image is well aligned with the training images. In
\cite{WagnerA2009-CVPR}, the framework of SBC was extended to iteratively align
a query image to one subject class. The face alignment algorithm in turn solves
the following minimization problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \tilde\bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain, which aligns $\tilde\bb$
with the training images from the $i$-th class $A_i$. The algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} to both to use the
$\ell_1$ norm as a robust error function, and also simultaneously 
to estimate the illumination model $A\x$.  
The original solution \cite{WagnerA2009-CVPR} was implemented based 
on standard interior-point
methods, and its speed has been since improved by ALM in a more recent paper
\cite{WagnerA2011-PAMI}. However, primarily due the high per-user cost
of the alignment step, the recognition system was still not fast enough
for real-time recognition against datasets of hundreds of users.
In this paper, in addition to solving the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
discuss how to properly accelerate the iterative face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, the face
recognition system has been fully implemented on a state-of-the-art
workstation, as shown in Figure \ref{fig:pipeline}. 
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:parallelism}.  We will show that a highly accurate and robust face
recognition system can be constructed to recognize face images from a 
database of several hundred subjects quickly enough to be practical for
access control applications. 
Note that for applications targeting thousands of users or more, it would likely be
necessary to distribute the work over a network of machines; this paper
focuses on the (shared memory) parallelism available in a single machine.
The C++ source code of our implementations, which target Nvidia's CUDA GPU architecture
will be made available for evaluation by other researchers.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{Overview of the face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

\subsection{Literature Review} 
In this section, we briefly review the literature of $\ell_1$-min and existing
works on its parallelization.  Traditionally, $\ell_1$-min (or the so-called
basis pursuit (BP)) has been formulated as a linear program
\cite{ChenS2001-SIAM}. Several variations of the solution are also well known
in optimization, including a noisy approximation via quadratic programming
called the LASSO \cite{TibshiraniR1996} and truncated Newton interior-point
method (TNIPM) \cite{KimS2007}.

One of the drawbacks of most interior-point methods for $\ell_1$-min (including
BP, LASSO, and TNIPM) is that they require the solution sequence to follow an
interior path
%{\bf DW: Following the interior path is what made them faster than
%traditional linear programming techniques that follow the edge of the feasible set;
%does Homotopy actually go outside the feasible set?  Just want to check if this is
%an accurate representation...}
, which is often computationally expensive. To mitigate these
issues, an approach called \emph{Homotopy} has been studied to accelerate the
speed of $\ell_1$-min. Homotopy methods for recovering sparse signals were
first studied in the context of LASSO \cite{OsborneM2000}, which inspired a
solution to the \emph{forward stagewise linear regression} problem called LARS
\cite{EfronB2004}, and eventually led to the corresponding Homotopy algorithms
for BP in \cite{MalioutovD2005,DonohoD2006}.

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves 
computationally expensive operations such as matrix-matrix multiplication and 
the solution of least-squares problems with varying $A$ matrices. 
In Section \ref{sec:ALM}, we contend
that ALM is a better choice among Homotopy and other existing $\ell_1$-min
solvers for implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) method \cite{WrightS2008,BeckA2009}.
IST algorithms mainly involves elementary operations such as vector 
algebra and matrix-vector multiplication. Therefore,
when the dimension of the problem becomes high, IST-type algorithms are
particularly suitable for systems in parallel-computing environments. In
\cite{YangA2010-ICIP}, the authors further showed that, compared to other
IST-type algorithms, ALM is able to achieve good estimation accuracy comparable
to BP and Homotopy, while at the same time significantly improve the speed
compared to these exact $\ell_1$-min solutions. Due to these facts, in this
paper, we choose ALM as the core algorithm to implement the complete parallel
face recognition pipeline.

In terms of the past works in parallel $\ell_1$-min, the literature has been
limited, to the best of our knowledge. In \cite{BorghiA2010}, Borghi et al.
developed a special proximal gradient $\ell_1$-min algorithm based on
Moreau-Yosida regularization. In \cite{MurphyM2010}, Murphy et al. presented
parallel implementation of the l1-SPIRIT MRI reconstruction algorithm on NVidia
programmable GPUs.

@@

Lagrange multiplier methods have been frequently used in convex programming to
eliminate equality constraints via adding a significant penalty term to the
cost function for infeasible points. ALM methods differ from other
penalty-based approaches by simultaneously estimating the optimal solution and
Lagrange multipliers in an iterative fashion.  For $\ell_1$-min, the augmented
Lagrange function is defined as: \begin{equation} L_\mu(\xx,\yy) = \|\xx\|_1 +
\yy^T(\bb - A\xx - \ee) + \frac{\mu}{2} \| \bb-A\xx-\ee \|_2^2, \end{equation}
where $\mu > 0$ is a constant that penalizes infeasibility and $\yy$ is a
vector of lagrange multipliers.

In Lagrange Multiplier Theory, if there exists a Lagrangian $y^*$ that
satisfies the second-order sufficiency conditions for optimality
\cite{BertsekasD2003}, then for a sufficiently large $mu$, the optimal
$\ell_1$-min solution also minimizes
\begin{equation}
\xx^* = \arg \min L_\mu(\xx,\yy^*).
\label{eq:optimal-ALM}
\end{equation}

In practice, the optimal values for the triplet $(\xx^*, \yy^*, \uu)$ are all
unknown. Furthermore, it has been observed that if solving
\eqref{eq:optimal-ALM} with a large initial value of $\mu$ tends to lead to
slower convergence speed \cite{WrightS2008,YangA2010-ICIP}. In
\cite{BertsekasD2003,YangJ2009}, an alternating procedure has been proposed to
iteratively update $\xx$ and $\yy$:
\begin{equation}
\left \{
\begin{array}{lll}
\xx_{k+1} & = & \arg\min_{\xx} \, L_{\mu_k} (\xx,\yy_k)\\
\yy_{k+1} & = & \yy_k + \mu_k (\bb - A\xx_{k+1}) \\
\end{array}
\right . ,
\label{eq:alm}
\end{equation}
where $\left\{\mu_{k}\right\}$ is a monotonically increasing positive sequence.
The iteration terminates when the estimates $(\xx, \yy)$ converges.

Note that in the iterative procedure outlined in \eqref{eq:alm}, the second
step only involves vector algebra and matrix-vector multiplication. Therefore,
the procedure is computationally efficient only if it is easier to minimize the
augmented Lagrange function compared to solving the original problem
\eqref{eq:l1min} directly. In fact, the first step that minimizes $L_{\mu_k}
(\xx,\yy_k)$ assuming $\yy_k$ and $\mu_k$ are fixed can be solved element-wise
iteratively by a soft-thresholding algorithm \cite{WrightS2008,BeckA2009},
whose time complexity is bounded by $O(n^2)$ and can be easily parallelized.
Algorithm \eqref{alg:alm} summarizes the ALM $\ell_1$-min algorithm \footnote{While we
present the ALM algorithm in the primal domain, there also
exists an implementation ALM in the dual domain.}.

 \begin{algorithm}[h]
\caption{Augmented Lagrange Multiplier (ALM)}
{\bf INPUT:} $\bb \in \Re^m$, $A \in \Re^{m \times n}$, $\tau\leftarrow$ largest eigenvalue of $A^TA$, and a constant $\rho>1$.
\begin{algorithmic}[1]
\WHILE{not converged ($k = 1,2,\ldots$)} \STATE $t_1 \leftarrow 1$, $\zz_1
\leftarrow \xx_k$, $\uu_1 \leftarrow \xx_k$ \WHILE{not
converged ($l = 1,2,\ldots$)} \STATE $\uu_{l+1}  \leftarrow
\mathrm{soft}\left(\zz_l - \frac{1}{\tau}A^T\left(A\zz_l - \bb
- \frac{1}{\mu_k}\yy_k\right), \frac{1}{\mu_k\tau}\right)$
\STATE $t_{l+1} \leftarrow \frac{1}{2}\left( 1 +
\sqrt{1+4t_l^2}\right)$ \STATE $\zz_{l+1} \leftarrow \uu_{l+1}
+ \frac{t_l - 1}{t_{l+1}}(\uu_{l+1} - \uu_l)$ \ENDWHILE \STATE$
\xx_{k+1} \leftarrow \uu_{l+1}$ \STATE $\yy_{k+1} \leftarrow
\yy_k + \mu_k (\bb - A\xx_{k+1})$ \STATE $\mu_{k+1} \leftarrow
\rho\cdot\mu_k$ \ENDWHILE 
\end{algorithmic}

@@

In this paper we will clearly distinguish between algorithm {\em parallelism},
which is a property of an algorithm, and hardware {\em concurrency}, which is a
property of a hardware architecture.  Parallelism in an algorithm provides the
opportunity to perform computations concurrently on hardware.  Algorithms often
exhibit multiple levels of parallelism, and hardware often provides multiple
levels of concurrency.  In these terms, the primary contribution of this paper
will be to determine the optimal mappings between the parallelism available in
$\ell_1$-Minimization and the concurrency available in multi-core CPU and GPU
architectures.

In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since on-chip
memory (cache) is often orders of magnitude faster than off-chip memory, the
sizes and speeds of on-chip memory (cache) have big performance implications.
We give a brief overview of the caches that are available in our target
architectures, and defer discussion of their performance implications to
sections \ref{sec:alignment_implementation_cpu} and
\ref{sec:alignment_implementation_gpu}.

We will primarily be comparing CPU and GPU architectures on a per-board basis:
for CPU implementation we make use of all the cores in as many CPUs are present
on one motherboard, and for GPU implementation we make use of a single
many-core Nvidia GPU chip on a PCI card. GPUs are commonly shipped with a
single GPU chip per PCI-express card.  Note that most engineering  workstations
are capable of supporting several GPU cards.  Unless otherwise specified, all
implementations utilize single precision floating point datatypes.

\subsection{CPU Hardware Concurrency} 
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
on-chip cache, and relatively high clock speeds.

The baseline architecture for our experiments is a Linux workstation with two
quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.  Each processor
has its own memory interface, and is directly interfaced to half of the RAM
installed in the machine.  The amount of RAM installed exceeds the amount used
by the algorithms, and is not an important performance consideration.

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figures/caches}
\end{center}
\caption{Scale Drawing of the CPU and GPU cache sizes}
\label{fig:caches}
\end{figure}

% CPU CACHE DISCUSSION
%2MB = 2**21 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**7  = 128 images.
Each core has a private 32\,KB L1 data cache and a private 256\,KB L2 cache.  
Between the 8 cores, this is a total of 2\,MB L2 cache.
For the low resolution images used in sparse representation based
face recognition (on the order of $64\times64$ pixel images), the L2 cache
can in principle hold up to 128 image vectors.
%16MB = 2**24 bytes = 16777216 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**10  = 1024 images.
Each processor further has 8MB of L3 cache that is shared by all four cores.
Overall, this means the algorithm has approximately 16MB of on-chip cache
available.  This corresponds to 1024 image vectors.  A scale drawing
of the caches of the processor we're targeting can be seen in Figure \ref{fig:caches}.

%With Hyperthreading enabled, each core can support two threads simultaneously.
For floating point instructions, each core also has a vector processing unit
(SSE) capable of performing the same arithmetic operation on four single
precision floating point values simultaneously.

@@

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
on-chip cache, and relatively low clock speeds.

Compared to CPU architectures, GPU architectures dedicate less of their transistors
to cache, and more of them for arithmetic logic. Our target GPU architecture
is the Nvidia Fermi architecture.  CUDA based GPUs are comprised of several 
streaming multiprocessors (SMP) (currently 14 to 16 for a high end GPU).

The GPU executes programs as SIMT (Single Instruction Multiple Threads) in
groups of 32 threads called \emph{warps}.  These warps are then grouped into
higher levels called \emph{thread blocks} and \emph{grids}.  It is impossible
to communicate between SMPs during gpu program execution, so computational tasks
are typically broken up into multiple gpu programs (kernels) that are executed
serially.

% GPU CACHE DISCUSSION
% for L2 cache
% 768KB = 786432 bytes
% 64*64*4 = 2**14 bytes = 16384 bytes per image
% 48 images
Each SMP contains its own L1-cache, which is divided between hardware managed
and software managed ("shared") memory.  Additionally, all SMPs share a common
L2-cache.  For our system, each SMP has 64 KB of L1 cache with a total of 896
KB of L1 cache/shared memory and 768KB of L2 cache. The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is balanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.  A scale drawing of the caches available on the GPU can be seen in Figure \ref{fig:caches}.

The GPU has its own memory system, any data the GPU uses must first be
transferred from CPU DRAM to GPU DRAM over PCI-Express.  For our application,
this transfer overhead is amortized over a large amount of computation and is
not a major concern.

Compared to a typical CPU, a GPU has much higher peak computational bandwidth
as well as data bandwith to DRAM.  However, since our application is primarily
of matrix-vector operations and image resampling operations, it has a very low
arithmetic intensity (i.e. we perform a very small number of arithmetic
operations for every number that has to be loaded), and thus performance is
typically constrained by memory (and cache) bandwidth.  At large problem sizes
the CPU is limited by its relatively low bandwidth to DRAM compared to the gpu,
while at small problem sizes the GPU is limited by the relatively high latency
of its access to DRAM.

\section{Parallelism in the Face Alignment Stage}
\label{sec:alignment}
There are essentially two levels of parallelism available for exploitation in
for iterative alignment in the face recognition application.  The primary goal
of this section is to introduce these two levels of parallelism.  Additionally,
in order for some important aspects of our implementations to be clear later,
we need to establish how the iterative alignment stage works in some
detail.

The inputs of the iterative alignment stage are the test image $f(x,y)$ stacked into an $M$-
dimensional vector $\f$, and the aligned and stacked training images $A_i$ for
each training user $i$. 
The algorithm repeatedly solves a linearized version of the minimization problem
in \eqref{eq:l1min_alignment}.  This involves two computational steps:
the computation of the linearization, and the solution of the minimization problem.

>>>>

\section{Introduction} 
$\ell_1$-minimization ($\ell_1$-min) has received much attention in recent
years due to important applications in compressive sensing
\cite{BrucksteinA2007} and sparse representation \cite{WrightJ2010-PIEEE}.  
$\ell_1$-min refers to finding the minimum $\ell_1$-norm solution to an
underdetermined linear system $\bb=A\xx$:
\begin{equation}
\min \|\xx\|_1\quad \mbox{ subj. to }\quad \bb = A\xx.
\label{eq:l1min}
\end{equation}
It is now well established that, under certain conditions
\cite{CandesE2005-IT_1,DonohoD2004}, the minimum $\ell_1$-norm solution is also
the \emph{sparsest} solution to the system \eqref{eq:l1min}.

@@

In this paper, we study parallelization of $\ell$-min on many-core CPU and GPU
architectures. Although $\ell_1$-min \eqref{eq:l1min} is a convex
program, conventional algorithms such as interior-point methods
\cite{ChenS2001-SIAM,TibshiraniR1996} are known to suffer from poor scalability
for large-scale applications such as face recognition. Recently, a number of
accelerated algorithms have been proposed that explicitly take advantage of
the special structure of $\ell_1$-min problems
\cite{LorisI2009,YangA2010-ICIP}. We investigate parallelization of a
state-of-the-art $\ell_1$-min solution based on the classical framework of
\emph{augmented Lagrangian methods} (ALM) \cite{BertsekasD2003,YangA2010-ICIP}.

While solving \eqref{eq:l1min_denoise} can achieve exceptional
recognition accuracy on public databases, its success also relies on the assumption
that the query image is well aligned with the training images. In
\cite{WagnerA2009-CVPR}, the framework of SBC was extended to iteratively align
a query image to one subject class. The face alignment algorithm in turn solves
the following minimization problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \tilde\bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain, which aligns $\tilde\bb$
with the training (frontal) images from the $i$-th class $A_i$. The algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} using the
$\ell_1$-norm as a robust error function.

Numerically, the original solution \cite{WagnerA2009-CVPR} was implemented based 
on standard interior-point methods, and its speed has been since improved by ALM in a more recent paper
\cite{WagnerA2011-PAMI}. However, due the high per-class cost
of the alignment step, the recognition system still fails to achieve \emph{real-time recognition} against datasets of hundreds or thousands of users.
In this paper, in addition to accelerating the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
also discuss how to properly accelerate the iterative face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

In this paper we will clearly distinguish between algorithm {\em parallelism},
which is a property of an algorithm, and hardware {\em concurrency}, which is a
property of a hardware architecture. Parallelism in an algorithm provides the
opportunity to perform computations concurrently on hardware.  Algorithms often
exhibit multiple levels of parallelism, and hardware often provides multiple
levels of concurrency.  In these terms, the primary contribution of this paper
will be to determine the optimal mappings between the parallelism available in
$\ell_1$-min and the concurrency available in multi-core CPU and GPU
architectures.

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a new face
recognition system has been fully implemented on a state-of-the-art
workstation for security and access control applications, as shown in Figure \ref{fig:pipeline}. 
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:parallelism}.
Finally, note that for applications targeting thousands of users or more, it would likely be
necessary to distribute the work over a network of computers, while this paper
focuses on the (shared memory) parallelism available in a single machine.
The C++ source code of our implementations, which target Nvidia's CUDA GPU architecture,
will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

\subsection{Literature Review} 
We first briefly review the literature of $\ell_1$-min and existing
works on its parallelization.  Traditionally, $\ell_1$-min (or the so-called
basis pursuit (BP)) has been formulated as a linear program
\cite{ChenS2001-SIAM}. Several variations of the solution are also well known
in optimization, including a noisy approximation via quadratic programming
called the LASSO \cite{TibshiraniR1996} and truncated Newton interior-point
method (TNIPM) \cite{KimS2007}.

One of the drawbacks of most interior-point methods for $\ell_1$-min (including
BP, LASSO, and TNIPM) is that they require the solution sequence to follow an
interior path via gradient descent or conjugate gradient methods, which are computationally expensive.
To mitigate these issues, an approach called \emph{Homotopy} has been recently studied to accelerate the
speed of $\ell_1$-min. Homotopy methods for recovering sparse signals were
first studied in the context of LASSO \cite{OsborneM2000}, which inspired a
solution to the \emph{forward stagewise linear regression} problem called LARS
\cite{EfronB2004}, and eventually led to the corresponding Homotopy algorithms
for BP in \cite{MalioutovD2005,DonohoD2006}.

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves 
computationally expensive operations such as matrix-matrix multiplication and 
the solution of least-squares problems with varying $A$ matrices. 
In Section \ref{sec:ALM}, we contend
that ALM is a better choice among existing $\ell_1$-min
solvers for implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.
IST algorithms mainly involve elementary operations such as vector 
algebra and matrix-vector multiplication. Therefore,
when the dimension of the problem becomes high, IST-type algorithms are
particularly suitable for systems with parallel-computing pipelines. In
\cite{YangA2010-ICIP}, the authors further showed that, compared to other
IST-type algorithms, ALM is able to achieve good estimation accuracy comparable
to BP and Homotopy, while at the same time significantly improve the speed
compared to these exact $\ell_1$-min solutions. Due to these facts, in this
paper, we choose ALM as the core algorithm to implement the complete parallel
face recognition pipeline.

In terms of the past works in parallel $\ell_1$-min, the literature has been
limited, to the best of our knowledge. In \cite{BorghiA2010}, Borghi et al.
developed a special proximal gradient $\ell_1$-min algorithm based on
Moreau-Yosida regularization. In \cite{MurphyM2010}, Murphy et al. presented
parallel implementation of the l1-SPIRIT MRI reconstruction algorithm on NVidia
programmable GPUs, which is similar to the hardware setup of this paper.

@@

Lagrange multiplier methods have been frequently used in convex programming to
eliminate equality constraints via adding a significant penalty term to the
cost function for infeasible points. ALM methods differ from other
penalty-based approaches by simultaneously estimating the optimal solution and
Lagrange multipliers in an iterative fashion.  For $\ell_1$-min, the augmented
Lagrangian is defined as: \begin{equation} L_\mu(\xx,\yy) = \|\xx\|_1 +
\yy^T(\bb - A\xx - \ee) + \frac{\mu}{2} \| \bb-A\xx-\ee \|_2^2, \end{equation}
where $\mu > 0$ is a constant that penalizes infeasibility and $\yy$ is a
vector of lagrange multipliers.

In Lagrange Multiplier Theory, if there exists a Lagrangian $y^*$ that
satisfies the second-order sufficiency conditions for optimality
\cite{BertsekasD2003}, then for a sufficiently large $\mu$, the optimal
$\ell_1$-min solution also minimizes
\begin{equation}
\xx^* = \arg \min L_\mu(\xx,\yy^*).
\label{eq:optimal-ALM}
\end{equation}

In practice, the optimal values for the triplet $(\xx^*, \yy^*, \mu)$ are all
unknown. Furthermore, it has been observed that solving
\eqref{eq:optimal-ALM} with a large initial value of $\mu$ tends to lead to
slower convergence speed \cite{WrightS2008,YangA2010-ICIP}. In
\cite{BertsekasD2003,YangJ2009}, an alternating procedure has been shown to
iteratively update $\xx$ and $\yy$:
\begin{equation}
\left \{
\begin{array}{lll}
\xx_{k+1} & = & \arg\min_{\xx} \, L_{\mu_k} (\xx,\yy_k)\\
\yy_{k+1} & = & \yy_k + \mu_k (\bb - A\xx_{k+1}) \\
\end{array}
\right . ,
\label{eq:alm}
\end{equation}
where $\mu_{k}\rightarrow \infty$ is a monotonically increasing positive sequence.
The iteration terminates when the estimates $(\xx, \yy)$ converge.

Note that in the iterative procedure \eqref{eq:alm}, the second
step only involves vector algebra and matrix-vector multiplication. Therefore,
the procedure is computationally efficient only if it is easier to minimize the
augmented Lagrangian $L_{\mu_k} (\xx,\yy_k)$ compared to solving the original problem
\eqref{eq:l1min} directly. In fact, this problem can be solved element-wise
iteratively by a soft-thresholding algorithm \cite{WrightS2008,BeckA2009},
whose time complexity is bounded by $O(n^2)$ and can be easily parallelized.
Algorithm \ref{alg:alm} summarizes the generic ALM $\ell_1$-min algorithm.\footnote{For conciseness, we
only present the ALM algorithm in the primal domain. There also
exist implementations in the dual domain \cite{YangJ2009,YangA2010-ICIP}.}

 \begin{algorithm}[h]
\caption{Augmented Lagrangian Method (ALM)}
{\bf INPUT:} $\bb \in \Re^m$, $A=[A_1,\cdots, A_K] \in \Re^{m \times n}$, $\tau\leftarrow \max\mbox{eig}(A^TA)$, and constant $\rho>1$.
\begin{algorithmic}[1]
\WHILE{not converged ($k = 1,2,\ldots$)} 
\STATE $t_1 \leftarrow 1$, $\zz_1 \leftarrow \xx_k$, $\uu_1 \leftarrow \xx_k$ 
\WHILE{not converged ($l = 1,2,\ldots$)} 
\STATE $\uu_{l+1}  \leftarrow \mbox{shrink}(\zz_l - \frac{1}{\tau}A^T(A\zz_l - \bb - \frac{1}{\mu_k}\yy_k), \frac{1}{\mu_k\tau})$
\STATE $t_{l+1} \leftarrow \frac{1}{2}( 1 + \sqrt{1+4t_l^2})$
\STATE $\zz_{l+1} \leftarrow \uu_{l+1}+ \frac{t_l - 1}{t_{l+1}}(\uu_{l+1} - \uu_l)$ 
\ENDWHILE 
\STATE $\xx_{k+1} \leftarrow \uu_{l+1}$ 
\STATE $\yy_{k+1} \leftarrow \yy_k + \mu_k (\bb - A\xx_{k+1})$ 
\STATE $\mu_{k+1} \leftarrow \rho\cdot\mu_k$ 
\ENDWHILE 
\end{algorithmic}

@@

In this section we discuss the levels of concurrency available in the hardware
architectures considered in this paper, as well as other aspects of the
hardware that are important for performance.  In particular, since on-chip
memory (cache) is often orders of magnitude faster than off-chip memory, the
size and speed of on-chip memory (cache) have big performance implications.
We give a brief overview of the caches that are available in our target
architectures, and defer discussion of their performance implications to
sections \ref{sec:alignment} and \ref{sec:recognition}.

We will primarily compare CPU and GPU architectures on a per-board basis:
for CPU implementation we make use of all the cores in as many CPUs are present
on one motherboard, and for GPU implementation we make use of a single
many-core Nvidia GPU chip on a PCI card. GPUs are commonly shipped with a
single GPU chip per PCI-express card.  Note that most engineering  workstations
are capable of supporting several GPU cards.  Unless otherwise specified, all
implementations utilize single precision floating point datatypes.

\subsection{CPU Hardware Concurrency} 
\label{sec:CPU-concurrency}
The main defining characteristics of contemporary multi-core CPU architectures
are that they have two levels of concurrency, relatively large amounts of
on-chip cache, and relatively high clock speed.

The baseline architecture for our experiment is a Linux workstation with two
quad-core Intel Nehalem E5530 processors clocked at 2.4 GHz.  Each processor
has its own memory interface, and is directly interfaced to half of the RAM
installed in the machine.  The amount of RAM installed exceeds the amount used
by the algorithms, and is not an important performance consideration.

% CPU CACHE DISCUSSION
%2MB = 2**21 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**7  = 128 images.
Each core has a private 32\,KB L1 data cache and a private 256\,KB L2 cache (\ie, a total of 2\,MB L2 cache available for the eight cores).
Therefore, for the low-resolution images used in face recognition (say, on the order of $64\times64$ grayscale images), the L2 cache
in principle can hold up to 16 images per core and 128 images in total.
%16MB = 2**24 bytes = 16777216 bytes
%64*64*4 = 2**14 bytes = 16384 bytes per image
%... so can fit 2**10  = 1024 images.
Each processor further has 8\,MB of L3 cache that is shared by the four cores.
Overall, the algorithm has approximately 16\,MB of L3 cache
available for a dual-processor configuration, which corresponds to about 1024 images.

%With Hyperthreading enabled, each core can support two threads simultaneously.
For floating-point instructions, each core also has a vector processing unit
(SSE) capable of performing the same arithmetic operation on four single-precision floating point values simultaneously.

@@

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
on-chip cache, and relatively low clock speed.

Compared to CPU architectures, modern GPU architectures dedicate less of their transistors
to cache, and more of them for arithmetic logic. Our target GPU architecture
is the Nvidia Fermi architecture (\eg, GTX 480 GPU used in this paper). CUDA based GPUs are comprised of several 
\emph{streaming multiprocessors} (SMP) (currently 14 to 16 for a high end GPU).

The GPU executes programs as SIMT (Single Instruction Multiple Threads) in
groups of 32 threads called \emph{warps}.  These warps are then grouped into
higher levels called \emph{thread blocks} and \emph{grids}.  Since SMPs are not designed
to communicate between themselves during GPU program execution, tasks
are typically broken up into multiple GPU programs (kernels) that are executed
serially.\footnote{Please add a footnote to further explain.}

Each SMP contains its own L1 cache, which is divided between hardware managed
and software managed (``shared'') memory.  Additionally, all SMPs share a common
L2 cache.  For our system, each SMP has 64 KB of L1 cache with a total of 896
KB of L1 cache/shared memory and 768KB of L2 cache. The relatively small amount
of cache (1/23 as much as CPU L3) on the GPU is ballanced by a significantly
higher bandwidth between the processor chip and off-chip memory (DRAM) compared
to the CPU.

Compared to typical CPUs, GPUs have much higher peak computational bandwidth
as well as data bandwidth to DRAM.  However, since our system primarily consists
of matrix-vector operations and image resampling operations, it has a very low
arithmetic intensity (\ie, we perform a very small number of arithmetic
operations for every number that has to be loaded), and thus performance is
typically constrained by memory (and cache) bandwidth.  At large problem sizes
the CPU is limited by its relatively low bandwidth to DRAM compared to the GPU,
while at small problem sizes the GPU is limited by the relatively high latency
of its access to DRAM.

\section{Parallelism in the Face Alignment Stage}
\label{sec:alignment}

Given the hardware concurrency in modern parallel computing architectures, next,
we shall study the optimal mappings of the algorithm parallelism in face recognition to the multi-core CPUs and GPUs, respectively.
In this section, we first focus on the face alignment stage (see Figure \ref{fig:pipeline}).

Face alignment \eqref{eq:l1min_alignment} estimates an image transformation $\tau$ that rectifies the query image $\tilde{\bb}$ with possible pose variation
w.r.t. each training class $A_i$, which leads to the minimal sparsity in error $\ee$ after the alignment. Note that directly solving \eqref{eq:l1min_alignment} would require
simultaneously estimating $\xx$, $\ee$, and $\tau$ typically on a nonlinear manifold, which renders the problem extremely difficult to solve exactly with many local minima.

To mitigate the problem, if the system can obtain to a good initial guess of the transformation $\tau$ (\eg, provided by a good face detector),
the optimal solution for $\tau$ can be sought iteratively by a sequence of linearized approximations $j=1, 2, \cdots$ as follows:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and $\Delta \tau$ is a step update to $\tau$. Denote $\bb_j = \tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta \tau^T]$, then the update $\Delta \tau$ can be computed by solving the following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
